Visual narratives of the COVID-19 pandemic

### 2022-10-31 JDSSV_manuscript_submitted.Rnw
FinalCopyedit
FinalFinalCopyedit

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{Adalbert F.X. Wilhelm\\
School for Business, Social and Decision Sciences\\
Jacobs University Bremen\\
Campus Ring 1\\
28759 Bremen\\
Germany\\
e-mail:a.wilhelm@jacobs-university.de
}

@@

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles. Hence data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and has already become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.

@@

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organisations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyse in this paper did not simply pop into being -- they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

@@

"The plague is always the others." This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of COVID-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

>>>>

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Susan Vanderplas\\
  Department of Statistics\\
  University of Nebraska–Lincoln\\
  349A Hardin Hall \\
  Lincoln, NE 68583-0963, USA\\
  E-mail: \email{susan.vanderplas@unl.edu}\\
  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
\And\\
  Adalbert F.X. Wilhelm\\
  School of Business, Social and Decision Sciences\\
  Jacobs University Bremen gGmbH\\
  Campus Ring 1\\
  28759 Bremen, Germany\\
  E-mail: \email{a.wilhelm@jacobs-university.de}\\
  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
}

@@

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles. Hence, data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and has already become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.

@@

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organisations, but may also be deliberately chosen for visual effect; colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyse in this paper did not simply pop into being -- they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

@@

"The plague is always the others" \citep{thiessen2021}. This is the short formula for dealing with infectious diseases from a historical perspective. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of COVID-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

### 2022-10-31 JDSSV_manuscript_submitted.tex
FinalFinalCopyedit

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{Adalbert F.X. Wilhelm\\
School for Business, Social and Decision Sciences\\
Jacobs University Bremen\\
Campus Ring 1\\
28759 Bremen\\
Germany\\
e-mail:a.wilhelm@jacobs-university.de
}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% \SweaveOpts{concordance=TRUE}

>>>>

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Susan Vanderplas\\
  Department of Statistics\\
  University of Nebraska–Lincoln\\
  349A Hardin Hall \\
  Lincoln, NE 68583-0963, USA\\
  E-mail: \email{susan.vanderplas@unl.edu}\\
  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
\And\\
  Adalbert F.X. Wilhelm\\
  School of Business, Social and Decision Sciences\\
  Jacobs University Bremen gGmbH\\
  Campus Ring 1\\
  28759 Bremen, Germany\\
  E-mail: \email{a.wilhelm@jacobs-university.de}\\
  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% \SweaveOpts{concordance=TRUE}
